The Blessing of FedAvg's Linear Speedup in Federated Learning

### 2020-06-01 source/sec_impact.tex
reorg exp
Update on Overleaf.

% Broader impact: Authors are asked to include a section in their submissions discussing the broader impact of their work, including possible societal consequences — -both positive and negative.
\section{Broader impact}
With the widespread use of mobile devices, the powerful sensors
collected enormous amount of data empowered with machine learning
solutions can transform people's life in many aspects. The proper
data integration can produce positive social
impact in terms of better personalization~\cite{fallah2020personalized}, more efficient information solicitation~\cite{chen2018federated}, more robust and stable service~\cite{47586}.

In the meantime, the increasing concerns of security and
privacy leaking from both large-companies and the entire society
present significant challenges of proper utilization of local
data. This trade-off between efficiency and privacy or security
lies in the center of many industrial applications, including 
but not limited to recommender systems~\cite{chen2018federated}, virtual assistants~\cite{lamautonomy}, and mobile keyboard prediction~\cite{47586}

In this work, we establish solid theoretical foundations for federated
learning that achieves efficiency and privacy at the same time. 
On one hand, the data is isolated in each local devices, thus prevent
the potential leaking of private information. On the otherhand, the
convergence rates improved substantially if the number of
active local devices increases. In conclusion, our work 
shows a faithful pathway to simultaneously secure people's privacy and improve
users' experience.

>>>>

% Broader impact: Authors are asked to include a section in their submissions discussing the broader impact of their work, including possible societal consequences — -both positive and negative.
\section{Broader impact}
With the widespread use of mobile devices, the powerful sensors
collected enormous amount of data empowered with machine learning
solutions can transform people's life in many aspects. The proper
data integration can produce positive social
impact in terms of better personalization~\cite{fallah2020personalized}, more efficient information solicitation~\cite{chen2018federated}, and enhanced service~\cite{47586}.

In the meantime, the increasing concerns of security and
privacy leaking from both large companies and the entire society
present significant challenges of proper utilization of local
data. This trade-off between efficiency and privacy/security
lies in the center of many industrial applications, including 
but not limited to recommendation systems~\cite{chen2018federated}, virtual assistants~\cite{lamautonomy}, and mobile keyboard prediction~\cite{47586}.

In this work, we establish solid theoretical foundations for federated
learning that achieves efficiency and privacy at the same time. 
On one hand, the data is isolated in each local devices, thus prevent
the potential leakage of private information. On the other hand, the
convergence rates can be improved substantially if the number of
active local devices increases. In conclusion, our work 
shows a faithful pathway to simultaneously secure people's privacy and improve
users' experience.

### 2020-06-01 source/sec_intro.tex
Update on Overleaf.

\section{Introduction}
% Motivate FL
The widespread usage of smart home devices and mobile phones constitutes
large-scale distributed networks that generate amount of personal data.
Models learned from such data lead to 
% The emergence of those massive amounts of data has empowered machine learning
% solutions to develop 
many successful applications such as personalized
recommendations~\cite{chen2018federated}, keyboard predictions~\cite{47586},
and etc. Due to the rising awareness of the privacy and rapid growth of
computational power in mobile devices, it becomes increasingly attractive to
push the computation to the edge. Recent years witnessed the rise of
federated learning~\cite{mcmahan2016communication} to fulfill this demand.

% Main challenge in FL.
As a distributed learning paradigm, Federated Learning (FL) collaboratively
learns a single statistical model leveraging the data from all devices while
keeping the training data isolated.  Thus, FL integrates cross-device
information for acquiring improved models without transferring training data from individual devices, thus improves user
privacy~\cite{kairouz2019advances}. Comparing to the traditional distribution optimization,  FL confronts unique
challenges~\cite{li2019convergence,smith2017federated}, including statistical heterogeneity: the data distributions in different devices can be significantly different, and system heterogeneity: only a subset of devices may be accessible to the central server for model training at each time.

% Gaps in the theoretical understanding. 
To address the aforementioned challenges, Federated
Averaging~\cite{mcmahan2016communication} (FedAvg) was proposed as an
effective heuristic and its great empirical success~\cite{47586} attracted
lots of efforts on building up its theoretical understanding 
and accelerated variants. 
Existing convergence guarantees of FedAvg were largely established on simplified setting~\cite{stich2018local,khaled2019first,wang2018cooperative,yu2019parallel} or provides a suboptimal 
convergence rate~\cite{li2019convergence}. Furthermore, the convergence of
FedAvg's accelerated variants~\cite{yu2019linear,huo2020faster,liu2019accelerating} under the realistic setting is yet to be analyzed.

% Our contributions
% \subsection{Contributions}
\textbf{Contributions.}
This work focus on providing comprehensive convergence analysis
of FedAvg and its accelerated variants under a realistic setting.
Our results show
that the less number of communication rounds is needed to converge to a
target accuracy and preserve the linear speedup w.r.t. participating devices at the same time.
More specifically, for strongly convex and smooth problems, we show 
that FedAvg admits a convergence rate of 
{\small{$\cO(1/NT)$}} convergence, and local steps can be as large as {\small{$\cO(\sqrt{T/N})$}}. For convex and smooth objectives, we show the rate of {\small{$\cO(1/\sqrt{NT})$}} and local steps can be as large as {\small{$\cO(\sqrt{T/N})$}}. 
Moreover, we show that FedAvg with Nesterov acceleration also enjoys the same rates and speedup. 
%The same conclusions are equally applied for Nesterov accelerated FedAvg. 
To the best of our knowledge, this is the first linear speedup results for FedAvg with Nesterov accelerated SGD. Furthermore, 
we provide the first improved convergence rate for Nesterov accelerated FedAvg comparing to FedAvg under the popular overparamterized setting. 
Last but not least, we systematically examine the convergence rate of
FedAvg and it accelerated variants on both a real-world dataset and a synthetic dataset. The empirical observations are well-aligned with theoretical indications.

@@

\textbf{Notations.}
Throughout the paper, we adopt the following notations. Let $N$ be the total
number of local devices, and $K \leq N$ is the maximal number of devices
that are accessible to the central server.  Let $T$ be the total number of  Stochastic Gradient Descent (SGD) iterations performed by each local device, $E$ be the local steps between two consecutive server communications. Thus, $T/E$ is
the number of communications.

>>>>

\section{Introduction}
% Motivate FL
The widespread usage of smart home devices and mobile phones constitutes
large-scale distributed networks that generate significant amount of personal data.
Models learned from such data have led to 
% The emergence of those massive amounts of data has empowered machine learning
% solutions to develop 
many successful applications such as personalized
recommendations~\cite{chen2018federated}, virtual assistants~\cite{lamautonomy}, and keyboard predictions~\cite{47586}, etc. Due to the rising awareness of  privacy and the rapid growth of
computational power in mobile devices, it becomes increasingly attractive to
push the computation to the edge. Recent years witnessed the emergence of
federated learning~\cite{mcmahan2016communication} to fulfill this demand.

% Main challenge in FL.
As a distributed learning paradigm, Federated Learning (FL) collaboratively
learns a single statistical model leveraging the data from a massive amount of devices while
keeping the training data isolated.  In particular, FL integrates cross-device
information to achieve improved model quality without transferring training data from individual devices, thus improves user
privacy~\cite{kairouz2019advances}. While promising, however, comparing to the traditional distribution optimization,  FL confronts unique
challenges~\cite{li2019convergence,smith2017federated}, including statistical heterogeneity, that is, the data distributions in different devices can be significantly different, and system heterogeneity, that is, only a subset of devices may be accessible to the central server for model training at each time.

% Gaps in the theoretical understanding. 
To address the aforementioned challenges, Federated
Averaging (FedAvg)~\cite{mcmahan2016communication} was proposed as an
effective heuristic which has enjoyed great empirical success~\cite{47586}, attracting
lots of efforts on building up its theoretical understanding 
and accelerated variants. 
Existing convergence guarantees of FedAvg were either largely established on simplified settings~\cite{stich2018local,khaled2019first,wang2018cooperative,yu2019parallel} or provides a suboptimal 
convergence rate~\cite{li2019convergence}. Furthermore, the convergence of
FedAvg's accelerated variants~\cite{yu2019linear,huo2020faster,liu2019accelerating} under the realistic setting is yet to be analyzed.

% Our contributions
% \subsection{Contributions}
\textbf{Contributions.}
This work focuses on providing comprehensive convergence analysis
of FedAvg and its accelerated variants under a realistic setting.
Our results show
that the less number of communication rounds is needed to converge to a
target accuracy while preserving the linear speedup w.r.t. the number of participating devices.
More specifically, for strongly convex and smooth problems, we show 
that FedAvg admits a convergence rate of 
{\small{$\cO(1/NT)$}}, and local steps can be as large as {\small{$\cO(\sqrt{T/N})$}}. For convex and smooth objectives, we show the rate of {\small{$\cO(1/\sqrt{NT})$}} and local steps can be as large as {\small{$\cO(\sqrt{T/N})$}}. 
Moreover, we show that FedAvg with Nesterov acceleration also enjoys the same rates and speedup. 
%The same conclusions are equally applied for Nesterov accelerated FedAvg. 
To the best of our knowledge, this is the first linear speedup results for FedAvg with Nesterov accelerated SGD. Furthermore, 
we provide the first improved convergence rate for Nesterov accelerated FedAvg comparing to FedAvg under the popular overparamterized setting. 
Last but not least, we systematically examine the convergence rate of
FedAvg and its accelerated variants on both a synthetic dataset and  a real-world dataset. The empirical observations are well-aligned with the obtained theoretical findings.

@@

\textbf{Notations.}
Throughout the paper, we adopt the following notations. Let $N$ be the total
number of local devices, and $K \leq N$ is the maximal number of devices
that are accessible to the central server.  Let $T$ be the total number of SGD iterations performed by each local device, $E$ be the local steps between two consecutive server communications. Thus, $T/E$ is
the number of communications.

### 2020-06-01 source/sec_relatedwork.tex
Update on Overleaf.

\section{Related Works}

@@

\textbf{The Convergence of FedAvg and Linear Speedup}
There has been a vast amount of convergence analysis for Federated averaging
(FedAvg) in different settings. However, most of the prior works provide the
convergence guarantee under the simplified setting, where either statistical
heterogeneity~\cite{stich2018local,zhou2017convergence,wang2018cooperative,woodworth2018graph} or system
heterogeneity~\cite{yu2019parallel,wang2019adaptive,khaled2019first,jiang2018linear} are not
considered. This simplified setting is also referred as local SGD. 
In~\cite{li2019convergence},
the authors firstly provide convergence of FedAvg for strongly convex problem considering both statistical heterogeneity and system heterogeneity, while their results
didn't achieve linear speedup. In~\cite{haddadpour2019convergence,liang2019variance,huo2020faster,jiang2018linear}, the convergence of FedAvg for non-convex problems has been studied. 
However, \cite{huo2020faster} did not discuss the linear speedup and the theorems in \cite{haddadpour2019convergence,liang2019variance} failed to cover the accelerated version of
FedAvg.

\textbf{Federated Learning Beyond Local SGD}
% The accelerated SGD methods has achieved optimal convergence rate in strongly convex
% optimization~\cite{nesterov1983method}.
In \cite{liu2019accelerating}, the authors provide a convergence rate of accelerated FedAvg for strongly convex problems, while it is considered in full participation setting.
\cite{yu2019linear} and \citep{huo2020faster} provide
convergence guarantee for Nesterov accelerated FedAvg for non-convex. 
However, \cite{yu2019linear} did not consider the system heterogeneity and \cite{huo2020faster} haven't discuss the linear speedup.
Furthermore, the accelerated FedAvg admitting a similar convergence rate as FedAvg in all previous works.  In this work, we provide a novel accelerated FedAvg with improved convergence rates under a popular overparameterized setting.

>>>>

\section{Related Work}

@@

\textbf{The Convergence of FedAvg and Linear Speedup.}
There has been a vast amount of convergence analysis for FedAvg in different settings. However, most of the prior works provide the
convergence guarantee under a simplified setting, where either statistical
heterogeneity~\cite{stich2018local,zhou2017convergence,wang2018cooperative,woodworth2018graph} or system
heterogeneity~\cite{yu2019parallel,wang2019adaptive,khaled2019first,jiang2018linear} are not
considered. This simplified setting is also referred as local SGD. 
In~\cite{li2019convergence},
the authors for the first time provide convergence of FedAvg for strongly convex problem considering both statistical heterogeneity and system heterogeneity, yet it
didn't achieve linear speedup. In~\cite{haddadpour2019convergence,liang2019variance,huo2020faster,jiang2018linear}, the convergence of FedAvg for non-convex problems has been studied. 
However, \cite{huo2020faster} did not discuss the linear speedup and the theorems in \cite{haddadpour2019convergence,liang2019variance} failed to cover the accelerated version of
FedAvg.

\textbf{Federated Learning Beyond Local SGD.}
% The accelerated SGD methods has achieved optimal convergence rate in strongly convex
% optimization~\cite{nesterov1983method}.
In \cite{liu2019accelerating}, the authors provide a convergence rate of accelerated FedAvg for strongly convex problems, while it is considered in full participation setting.
\cite{yu2019linear} and \citep{huo2020faster} provide
convergence guarantee for Nesterov accelerated FedAvg for non-convex. 
However, \cite{yu2019linear} did not consider the system heterogeneity and \cite{huo2020faster} haven't discuss the linear speedup.
Furthermore, the accelerated FedAvg admitting a similar convergence rate as FedAvg in all previous works.  In this work, we provide a novel accelerated FedAvg with improved convergence rates under a popular overparameterized setting.
